Fitting metabolic models to dissolved oxygen data: The Estuarine BAyesian Single-station Estimation method

### 2023-07-27 manu-draft.qmd
more edits from rn

For @eq-ebase, the terms on the right side of the equation are gross primary production (P, $aPAR$), ecosystem respiration (R), and gas exchange (D, remainder), respectively, processes that result in the change in dissolved oxygen ($dC$) per unit time ($dt$); $H$ is the water depth. Production is modeled linearly as a function of surface irradiance ($PAR$, $\text{W}/\text{m}^2$) and $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$). Respiration is modeled with the single parameter $R$ ($\text{mmol}/\text{m}^2/\text{d}$). Gas exchange is modeled using the formulation in @Wanninkhof14 as a wind-based parameterization for the gas transfer velocity [@Ho16] multiplied by the oxygen concentration gradient as the difference between the concentration measured in the surface water layer, $C$ ($\text{mmol}/\text{m}^3$) , and the saturation (or equilibrium) concentration, $C_{Sat}$ ($\text{mmol}/\text{m}^3$) calculated from salinity (psu) and temperature ($^\circ$C) [@Millero81; @Garcia92]. The gas transfer velocity, $k_w$ in @eq-kw, is modeled using wind speed squared at 10 meters above the water ($U_{10}^2$, $\text{m}^2/\text{s}^2$) and the Schmidt number ($Sc$), which is the ratio of the kinematic viscosity of water to the molecular diffusivity of oxygen in water [$S_c$ is computed from water temperature and salinity using the polynomial fit in @Wanninkhof14]. The $b$ parameter is estimated from the data and has units of $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$. As such, the three parameters estimated by EBASE are $a$, $R$, and $b$ using the required input data that includes time series of dissolved oxygen, water temperature, salinity, PAR, wind speed, and water column depth (`r run_reference(id = "tbl-ebaseiotab")`).

The EBASE approach was based on several modifications to BAyesian Single-station Estimation (BASE) the mass balance equation described in @Grace15:

$$
\frac{dC}{dt} = aPAR^p - r\Theta^{T - \bar{T}} + \left[K\cdot1.0241^{T-\bar{T}}\left(C_{sat} - C \right )\right]
$$ {#eq-base}

@@

A common theory behind most gas exchange parameterizations for metabolism estimates is that the transfer of gases of relatively low solubility, like oxygen, is proportional to the concentration gradient in a thin film of water at the air--water interface. It is also assumed that the water at the top of the film is saturated with respect to the atmosphere, so that the concentration difference across the thin film can be approximated as the difference between oxygen concentration measured in the surface water layer, $C$, and the saturation (or equilibrium) concentration $C_{Sat}$, which is usually calculated as a function of water temperature and salinity (the negative sign is needed because positive flux is set to be from water to air):

$$
D = -k_w \left(C_{sat}-C \right).
$$ {#eq-gasex}

The gas transfer velocity, $k_w$, has dimensions of length over time and increases as the water column in the vicinity of the air--water interface becomes more turbulent. As such, $k_w$ is usually modeled as a function of measurable variables that can potentially predict turbulent mixing of the surface water, e.g., wind speed, current velocity, and water depth [@Ho16]. $k_w$ also increases with temperature because the film of water at the air-sea interface becomes more thin as the random motion of gas and water molecules increases at higher temperatures and transfer of molecules across the film increases. This dependence of $k_w$ on temperature is usually modeled with the Schmidt number as in @eq-ebase, which combines information on the viscosity of the fluid and the diffusivity of the gas.

@@

where $\bar{T}$ is the average water temperature over each 24-hour period in the oxygen time series and $K$ is the reaeration coefficient estimated from the data. @eq-gasexbase indicates that $k_w$ increases with temperature and depth. The temperature dependence is consistent with the known decrease of $k_w$ with the Schmidt number, but the depth dependence is difficult to justify. An expectation is that $k_w$ would increase as water depth decreases for a given current speed because turbulence generated at the sediment--water interface by currents would more easily reach the surface [@Ho16]. Based on these differences, the wind-based parameterization for gas exchange was used for EBASE [@Wanninkhof14]. Additionally, $C_{sat}$ was calculated using only temperature and salinity, removing the need for barometric pressure used in BASE.

The additional modifications to @eq-base for EBASE were removing the $p$ and $\Theta$ parameters from the production (P) and respiration (R) components, respectively. The BASE method provides options to estimate both $\Theta$ and $p$ from the data or to set as fixed values defined in the model file ($\Theta = 1.07177$, $p=1$). For EBASE, the temperature dependency of respiration is removed by setting $\Theta^{(T-\bar{T})}$ = 1 and it is assumed that respiration is constant as in other methods [@Odum56; @Murrell18]. The $p$ parameter, which is dimensionless, is also set to 1 as in @Holtgrieve10, such that a linear relationship of production to $PAR$ was used. This removes the exponential dependence of production with increasing $PAR$. Removing these components also improved model stability by creating more reasonable estimates for the other parameters.

@@

The second advantage of EBASE that makes the model more appropriate for estuarine settings included several modifications to the underlying mass-balance equation for dissolved oxygen (@eq-ebase). As described in @sec-basemodif, EBASE is an estuarine adaptation of BASE (@eq-base) that was developed for stream applications and many assumptions of the underlying model of the latter were inappropriate for the former. Specifically, EBASE was modified to address three primary concerns of using BASE in an estuarine setting. First, the estimation of gross primary production as a function of PAR assumed a linear relationship such that the increase of production per unit increase of PAR was consistent for any level of the latter. The BASE model includes an exponential parameter, $p$, that defines a non-linear relationship between P and PAR (although it should be noted that BASE provides an option to set $p = 1$). @Holtgrieve10 describes a simple linear photosynthesis-irradiance curve as appropriate for systems that are never light-saturated and this assumption likely holds true for estuaries that are commonly more turbid than open-ocean or freshwater locations. Second, EBASE removed the temperature-dependency of respiration and assumed that respiration did not change with water temperature, consistent with applications of the Odum method to other shallow estuaries [@Beck15; @Murrell18]. This modification addresses recent work that suggests respiration in streams may be reduced during the night with lower temperatures as the respiration of labile autochthonous organic matter may contribute to changes in the dissolved oxygen time series [@sadro2014; @schindler2017]. There is currently no justification for this assumption in estuaries. Finally, the gas exchange in EBASE used the parameterization in @Wanninkhof14 that is primarily wind-based and developed for the open-ocean, which is clearly more appropriate than the prior formulation in BASE that includes a temperature and depth dependency for streams. Overall, these changes also reduced the required inputs and many monitoring locations nationwide include the requisite data for applying EBASE (i.e., dissolved oxygen, water temperature, salinity, PAR, and wind speed, as collected by the NERRS).

>>>>

For @eq-ebase, the terms on the right side of the equation are gross primary production (P, $aPAR$), ecosystem respiration (R), and gas exchange (D, remainder), respectively, processes that result in the change in dissolved oxygen ($dC$) per unit time ($dt$); $H$ is the water depth. Production is modeled linearly as a function of surface irradiance (PAR, $\text{W}/\text{m}^2$) and $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$). Respiration is modeled with the single parameter $R$ ($\text{mmol}/\text{m}^2/\text{d}$). Gas exchange is modeled using the formulation in @Wanninkhof14 as a wind-based parameterization for the gas transfer velocity [@Ho16] multiplied by the oxygen concentration gradient as the difference between the concentration measured in the surface water layer, $C$ ($\text{mmol}/\text{m}^3$) , and the saturation (or equilibrium) concentration, $C_{Sat}$ ($\text{mmol}/\text{m}^3$) calculated from salinity (psu) and temperature ($^\circ$C) [@Millero81; @Garcia92]. The gas transfer velocity, $k_w$ in @eq-kw, is modeled using wind speed squared at 10 meters above the water ($U_{10}^2$, $\text{m}^2/\text{s}^2$) and the Schmidt number ($Sc$), which is the ratio of the kinematic viscosity of water to the molecular diffusivity of oxygen in water [$S_c$ is computed from water temperature and salinity using the polynomial fit in @Wanninkhof14]. The $b$ parameter is estimated from the data and has units of $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$. As such, the three parameters estimated by EBASE are $a$, $R$, and $b$ using the required input data that includes time series of dissolved oxygen, water temperature, salinity, PAR, wind speed, and water column depth (`r run_reference(id = "tbl-ebaseiotab")`).

@@

@eq-ebase is similar to the oxygen mass balance equation in BASE [@Grace15], with the main difference being the treatment of gas exchange. The gas transfer velocity, $k_w$, increases as the water column in the vicinity of the air--water interface becomes more turbulent. As such, $k_w$ is usually modeled as a function of measurable variables that can potentially predict turbulent mixing of the surface water, e.g., wind speed, current velocity, and water depth [@Ho16]; fetch and suspended solids have also been considered for estuarine systems [@Borges11]. $k_w$ also increases with temperature because the film of water at the air-sea interface becomes thinner as the viscosity decreases and the random motion of gas molecules increases. This dependence of $k_w$ on temperature is usually modeled with the Schmidt number as in @eq-ebase, which combines information on the viscosity of the fluid and the diffusivity of the gas.

@@

where $\bar{T}$ is the average water temperature over each 24-hour period in the oxygen time series and $K$ is the reaeration coefficient estimated from the data. @eq-gasexbase indicates that $k_w$ increases with temperature and depth. The temperature dependence is consistent with the known decrease of $k_w$ with the Schmidt number, but the depth dependence is difficult to justify. An expectation is that $k_w$ would increase as water depth decreases for a given current speed because turbulence generated at the sediment--water interface by currents would more easily reach the surface [@Ho16]. Based on these differences and the lack of availability of other possible drivers of the transfer velocity in coastal systems (e.g., currents, fetch, suspended solids, and wave height), the wind-based parameterization for gas exchange was used for EBASE [@Wanninkhof14]. However, this parameterization is largely based on open-ocean data and so we allow $b$, which was esimated by @Wanninkhof14 to be 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$, to be a free parameter that may reflect spatial and temporal variability in factors other than wind and temperature that may influence transfer velocity.

BASE was also modified by removing the temperature dependence of respiration and the potential for $P$ to depend nonlinearly on PAR. While there is evidence for both dependencies in aquatic systems, we found, in preliminary model runs, that model stability was enhanced when these dependencies were removed. It would be straightforward to include such dependencies in EBASE as our understanding of them improve and as data become available that will allow them to be robustly constrained.

@@

The second advantage of EBASE that makes the model more appropriate for estuarine settings included several modifications to the underlying mass-balance equation for dissolved oxygen (@eq-ebase). As described in @sec-basemodif, EBASE is an estuarine adaptation of BASE [@Grace15] that was developed for stream applications and many assumptions of the underlying model of the latter were inappropriate for the former. Specifically, EBASE was modified to address three primary concerns of using BASE in an estuarine setting. First, the estimation of gross primary production as a function of PAR assumed a linear relationship such that the increase of production per unit increase of PAR was consistent for any level of the latter. The BASE model includes an exponential parameter, $p$, that defines a non-linear relationship between P and PAR (although it should be noted that BASE provides an option to set $p = 1$). @Holtgrieve10 describes a simple linear photosynthesis-irradiance curve as appropriate for systems that are never light-saturated and this assumption likely holds true for estuaries that are commonly more turbid than open-ocean or freshwater locations. Second, EBASE removed the temperature-dependency of respiration and assumed that respiration did not change with water temperature, consistent with applications of the Odum method to other shallow estuaries [@Beck15; @Murrell18]. This modification addresses recent work that suggests respiration in streams may be reduced during the night with lower temperatures as the respiration of labile autochthonous organic matter may contribute to changes in the dissolved oxygen time series [@sadro2014; @schindler2017]. There is currently no justification for this assumption in estuaries. Finally, the gas exchange in EBASE used the parameterization in @Wanninkhof14 that is primarily wind-based and developed for the open-ocean, which is clearly more appropriate than the prior formulation in BASE that includes a temperature and depth dependency for streams. Overall, these changes also reduced the required inputs and many monitoring locations nationwide include the requisite data for applying EBASE (i.e., dissolved oxygen, water temperature, salinity, PAR, and wind speed, as collected by the NERRS).
